---
title: "ARIMA Modeling"
author: "Mohammed Ali"
date: "March 2, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(astsa)
library(xts)
```

# Time Series Data and Models

Investigate the nature of time series data and learn the basics of ARMA models that can explain the behavior of such data. we will learn the basic R commands needed to help set up raw time series data to a form that can be analyzed using ARMA models.

## Data Play
Before we use a data set for the first time, we should use the help system to see the details of the data. For example, use `help(AirPassengers)` or `?AirPassengers` to see the details of the series.

```{r data_play}
# View a detailed description of AirPassengers
#help(AirPassengers)

# Plot AirPassengers
plot(AirPassengers)

# Plot the DJIA daily closings
plot(djia$Close)

# Plot the Southern Oscillation Index
plot(soi)


```


## Differencing
When a time series is _trend stationary_, it will have stationary behavior around a trend. A simple example is $Y_t=α+β_t+X_t$ where $X_t$ is stationary.

A different type of model for trend is _random walk_, which has the form $X_t=X_{t−1}+W_t$, where $W_t$ is white noise. It is called a random walk because at time t the process is where it was at time $t−1$ plus a completely random movement. For a random walk with drift, a constant is added to the model and will cause the random walk to drift in the direction (positve or negative) of the drift.


In both cases, simple differencing can remove the trend and coerce the data to stationarity. Differencing looks at the difference between the value of a time series at a certain point in time and its preceding value. That is, $X_t−X_{t−1}$ is computed.

To check that it works, we will difference each generated time series and plot the detrended series. If a time series is in x, then `diff(x)` will have the detrended series obtained by differencing the data. To plot the detrended series, simply use `plot(diff(x))`.

```{r diff}
# Plot globtemp and detrended globtemp
par(mfrow = c(2,1))
plot(globtemp) 
plot(diff(globtemp)) 

# Plot cmort and detrended cmort
par(mfrow = c(2,1))
plot(cmort)
plot(diff(cmort))

```

## Dealing with Trend and Heteroscedasticity
Here, we will coerce nonstationary data to stationarity by calculating the return or growth rate as follows.

Often time series are generated as
$X_t=(1+p_t)X_{t−1}$
meaning that the value of the time series observed at time t equals the value observed at time $t−1$ and a small percent change pt at time $t$.

A simple deterministic example is putting money into a bank with a fixed interest $p$. In this case, $X_t$ is the value of the account at time period t with an initial deposit of $X_0$.

Typically, pt is referred to as the return or growth rate of a time series, and this process is often stable.

For reasons that are outside the scope of this course, it can be shown that the growth rate pt can be approximated by
$Y_t=logX_t−logX_t−1≈p_t$.

```{r diff_log}
# astsa and xts are preloaded 

# Plot GNP series (gnp) and its growth rate
par(mfrow = c(2,1))
plot(gnp)
plot(diff(log(gnp)))

# Plot DJIA closings (djia$Close) and its returns
par(mfrow = c(2,1))
plot(djia$Close)
plot(diff(log(djia$Close)))
```

## Simulating ARMA Models
Any stationary time series can be written as a linear combination of white noise. In addition, any ARMA model has this form, so it is a good choice for modeling stationary time series.

R provides a simple function called `arima.sim()` to generate data from an ARMA model. For example, the syntax for generating 100 observations from an `MA(1)` with parameter `.9` is `arima.sim(model = list(order = c(0, 0, 1), ma = .9 ), n = 100`). We can also use `order = c(0, 0, 0)` to generate white noise.

Now, we will generate data from various ARMA models. For each command, generate 200 observations and plot the result.
```{r arma_model}
# Generate and plot white noise
WN <- arima.sim(model = list(order = c(0, 0, 0)), n = 200)
plot(WN)


# Generate and plot an MA(1) with parameter .9 
MA <- arima.sim(model = list(order = c(0, 0, 1), ma = 0.9), n = 200)
plot(MA)


# Generate and plot an AR(2) with parameters 1.5 and -.75
AR <- arima.sim(model = list(order = c(2, 0, 0), ar = c(1.5, -0.75)), n = 200)
plot(AR)
```

# Fitting ARMA models
Discover the wonderful world of ARMA models and how to fit these models to time series data. We will learn how to identify a model, how to choose the correct model, and how to verify a model once we fit it to data. We will learn how to use R time series commands from the stats and astsa packages.

## Fitting an AR(1) Model

```{r ar_1}
# Generate 100 observations from the AR(1) model
x <- arima.sim(model = list(order = c(1, 0, 0), ar = .9), n = 100) 

# Plot the generated data 
plot(x)

# Plot the sample P/ACF pair
acf2(x)

# Fit an AR(1) to the data and examine the t-table
sarima(x, p = 1, d = 0, q = 0)
```

## Fitting an AR(2) Model
```{r ar_2}
# Plot x
x <- arima.sim(model = list(order = c(2, 0, 0), ar = c(1.5, -.75)), n = 200)
plot(x)

# Plot the sample P/ACF of x
acf2(x)

# Fit an AR(2) to the data and examine the t-table
sarima(x, p = 2, d = 0, q = 0)
```

## Fitting an MA(1) Model
```{r ma_1}
# Plot x
x <- arima.sim(model = list(order = c(0, 0, 1), ma = -.8), n = 100)
plot(x)
# Plot the sample P/ACF of x
acf2(x)

# Fit an MA(1) to the data and examine the t-table
sarima(x, p = 0, d = 0, q = 1)
```

## Fitting an ARMA model
```{r arma_2_1}
x <- arima.sim(model = list(order = c(2, 0, 1), ar = c(1, -.9), ma = .8), n = 250)

# Plot x
plot(x)

# Plot the sample P/ACF of x
acf2(x)

# Fit an ARMA(2,1) to the data and examine the t-table
sarima(x, p = 2, d = 0, q = 1)
```

## Model Choice - I
```{r model_1}
dl_varve <- diff(log(varve))

# Fit an MA(1) to dl_varve.   
sarima(dl_varve, p = 0, d = 0, q = 1)

# Fit an MA(2) to dl_varve. Improvement?
sarima(dl_varve, p = 0, d = 0, q =2)

# Fit an ARMA(1,1) to dl_varve. Improvement?
sarima(dl_varve, p = 1, d = 0, q = 1)
```

## ARMA get in
The data in oil are crude `oil`, WTI spot price FOB (in dollars per barrel), weekly data from 2000 to 2008. Let us use our skills to fit an ARMA model to the returns. The weekly crude oil prices (oil) are plotted for us. Throughout the exercise, work with the returns, which we will calculate.

As before, the astsa package is preloaded for we. The data are preloaded as oil and plotted on the right.

```{r wrap}
# Calculate approximate oil returns
oil_returns <- diff(log(oil))

# Plot oil_returns. Notice the outliers.
plot(oil_returns)

# Plot the P/ACF pair for oil_returns
acf2(oil_returns)

# Assuming both P/ACF are tailing, fit a model to oil_returns
sarima(oil_returns, p = 1, d = 0, q = 1)
```


# ARIMA Models
Now that we know how to fit ARMA models to stationary time series, we will learn about integrated ARMA (ARIMA) models for nonstationary time series. we will fit the models to real data using R time series commands from the stats and astsa packages.

## ARIMA - Plug and Play
```{r arima_1}
x <- arima.sim(model = list(order = c(1, 1, 0), ar = .9), n = 200)
# Plot x
plot(x)

# Plot the P/ACF pair of x
acf2(x)

# Plot the differenced data
plot(diff(x))

# Plot the P/ACF pair of the differenced data
acf2(diff(x))
```

## Global Warming
```{r global_Warming}
# Plot the sample P/ACF pair of the differenced data 
acf2(diff(globtemp))

# Fit an ARIMA(1,1,1) model to globtemp
sarima(globtemp, p = 1, d = 1, q = 1)

# Fit an ARIMA(0,1,2) model to globtemp. Which model is better?

sarima(globtemp, p = 0, d = 1, q = 2)
```

## Forecasting Simulated ARIMA
```{r forcast}
x <- arima.sim(model = list(order = c(1, 1, 0), ar = .9), n = 200) 
# Plot P/ACF pair of differenced data 
acf2(diff(x))

# Fit model - check t-table and diagnostics
sarima(x, p = 1, d = 1, q = 0)

# Forecast the data 20 time periods ahead
sarima.for(x, n.ahead = 20, p = 1, d = 1, q = 0) 
#lines(y)  
```

## Forecasting Global Temperatures
Now you can try forecasting real data.

Here, you will forecast the annual global temperature deviations globtemp to 2050. Recall that in previous exercises, you fit an `ARIMA(0,1,2)` model to the data. You will refit the model to confirm it, and then forecast the series 35 years into the future.

```{r forcast_globalTemp}
# Fit an ARIMA(0,1,2) to globtemp and check the fit
sarima(globtemp, p = 0, d = 1, q = 2)

# Forecast data 35 years into the future
sarima.for(globtemp, n.ahead = 35, p = 0, d = 1, q = 2) 
```





# Seasonal ARIMA
We will learn how to fit and forecast seasonal time series data using seasonal ARIMA models. This is accomplished using what you learned in the previous chapters and by learning how to extend the R time series commands available in the stats and astsa packages.

## Data Analysis - Unemployment
```{r fit_pure_Seasonal}
# Plot unemp 
plot(unemp)

# Difference your data and plot it
d_unemp <- diff(unemp)
plot(d_unemp)

# Seasonally difference d_unemp and plot it
dd_unemp <- diff(d_unemp, lag = 12)
plot(dd_unemp)

# Plot P/ACF pair of fully differenced data to lag 60
acf2(dd_unemp)

# Plot P/ACF pair of fully differenced data to lag 60
dd_unemp <- diff(diff(unemp), lag = 12)


# Fit an appropriate model

acf2(dd_unemp, max.lag = 60)

# Fit an appropriate model
sarima(unemp, p = 2, d = 1, q = 0, P = 0, D = 1, Q = 1, S = 12)
```

## Data Analysis - Commodity Prices
```{r commodity_prices}
# Plot differenced chicken
plot(diff(chicken))

# Plot P/ACF pair of differenced data to lag 60
acf2(diff(chicken), max.lag=60)

# Fit ARIMA(2,1,0) to chicken - not so good
sarima(chicken, p = 2, d = 1, q = 0)

# Fit SARIMA(2,1,0,1,0,0,12) to chicken - that works
sarima(chicken, p = 2, d = 1, q = 0, P = 1, D = 0, Q = 0, S = 12)
```

## Data Analysis - Birth Rate
```{r birth_date}
# Plot P/ACF to lag 60 of differenced data
d_birth <- diff(birth)
acf2(d_birth, max.lag=60)

# Plot P/ACF to lag 60 of seasonal differenced data
dd_birth <- diff(d_birth, lag = 12)
acf2(dd_birth, max.lag=60)

# Fit SARIMA(0,1,1)x(0,1,1)_12. What happens?
sarima(birth, p = 0, d = 1, q = 1, P = 0, D = 1, Q = 1, S = 12)

# Add AR term and conclude

sarima(birth, p = 1, d = 1, q = 1, P = 0, D = 1, Q = 1, S = 12)
```

## Forecasting Monthly Unemployment
```{r unemployment_forcast}
# Fit your previous model to unemp and check the diagnostics
sarima(unemp, p = 2, d = 1, q = 0, P = 0, D = 1, Q = 1, S = 12)

# Forecast the data 3 years into the future

sarima.for(unemp,n.ahead = 36, p = 2, d = 1, q = 0, P = 0, D = 1, Q = 1, S = 12)
```

## How Hard is it to Forecast Commodity Prices?
```{r commodity_forcast}
# Fit the chicken model again and check diagnostics
sarima(chicken, p = 2, d = 1, q = 0, P = 1, D = 0, Q = 0, S = 12)

# Forecast the chicken data 5 years into the future
sarima.for(chicken,n.ahead = 60,  p = 2, d = 1, q = 0, P = 1, D = 0, Q = 0, S = 12)
```
