---
title: "ARIMA Modeling"
author: "Mohammed Ali"
date: "March 2, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(astsa)
library(xts)
```

# Time Series Data and Models

Investigate the nature of time series data and learn the basics of ARMA models that can explain the behavior of such data. we will learn the basic R commands needed to help set up raw time series data to a form that can be analyzed using ARMA models.

## Data Play
Before we use a data set for the first time, we should use the help system to see the details of the data. For example, use `help(AirPassengers)` or `?AirPassengers` to see the details of the series.

```{r data_play}
# View a detailed description of AirPassengers
#help(AirPassengers)

# Plot AirPassengers
plot(AirPassengers)

# Plot the DJIA daily closings
plot(djia$Close)

# Plot the Southern Oscillation Index
plot(soi)


```


## Differencing
When a time series is _trend stationary_, it will have stationary behavior around a trend. A simple example is $Y_t=α+β_t+X_t$ where $X_t$ is stationary.

A different type of model for trend is _random walk_, which has the form $X_t=X_{t−1}+W_t$, where $W_t$ is white noise. It is called a random walk because at time t the process is where it was at time $t−1$ plus a completely random movement. For a random walk with drift, a constant is added to the model and will cause the random walk to drift in the direction (positve or negative) of the drift.


In both cases, simple differencing can remove the trend and coerce the data to stationarity. Differencing looks at the difference between the value of a time series at a certain point in time and its preceding value. That is, $X_t−X_{t−1}$ is computed.

To check that it works, we will difference each generated time series and plot the detrended series. If a time series is in x, then `diff(x)` will have the detrended series obtained by differencing the data. To plot the detrended series, simply use `plot(diff(x))`.

```{r diff}
# Plot globtemp and detrended globtemp
par(mfrow = c(2,1))
plot(globtemp) 
plot(diff(globtemp)) 

# Plot cmort and detrended cmort
par(mfrow = c(2,1))
plot(cmort)
plot(diff(cmort))

```

## Dealing with Trend and Heteroscedasticity
Here, we will coerce nonstationary data to stationarity by calculating the return or growth rate as follows.

Often time series are generated as
$X_t=(1+p_t)X_{t−1}$
meaning that the value of the time series observed at time t equals the value observed at time $t−1$ and a small percent change pt at time $t$.

A simple deterministic example is putting money into a bank with a fixed interest $p$. In this case, $X_t$ is the value of the account at time period t with an initial deposit of $X_0$.

Typically, pt is referred to as the return or growth rate of a time series, and this process is often stable.

For reasons that are outside the scope of this course, it can be shown that the growth rate pt can be approximated by
$Y_t=logX_t−logX_t−1≈p_t$.

```{r diff_log}
# astsa and xts are preloaded 

# Plot GNP series (gnp) and its growth rate
par(mfrow = c(2,1))
plot(gnp)
plot(diff(log(gnp)))

# Plot DJIA closings (djia$Close) and its returns
par(mfrow = c(2,1))
plot(djia$Close)
plot(diff(log(djia$Close)))
```

# Simulating ARMA Models
Any stationary time series can be written as a linear combination of white noise. In addition, any ARMA model has this form, so it is a good choice for modeling stationary time series.

R provides a simple function called `arima.sim()` to generate data from an ARMA model. For example, the syntax for generating 100 observations from an `MA(1)` with parameter `.9` is `arima.sim(model = list(order = c(0, 0, 1), ma = .9 ), n = 100`). We can also use `order = c(0, 0, 0)` to generate white noise.

Now, we will generate data from various ARMA models. For each command, generate 200 observations and plot the result.
```{r arma_model}
# Generate and plot white noise
WN <- arima.sim(model = list(order = c(0, 0, 0)), n = 200)
plot(WN)


# Generate and plot an MA(1) with parameter .9 
MA <- arima.sim(model = list(order = c(0, 0, 1), ma = 0.9), n = 200)
plot(MA)


# Generate and plot an AR(2) with parameters 1.5 and -.75
AR <- arima.sim(model = list(order = c(2, 0, 0), ar = c(1.5, -0.75)), n = 200)
plot(AR)
```


